---
title: "Lecture Assignment 5"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```


\newpage
**Part 4.4**

**Question 1** 

The code does not work because "my_varıable" is missing a dot on the 'i'. Therefore, "my_varıable" is not found because it was never created as an object. Instead, "my_variable" was created as an object. This is a small syntax error. 

```{r Q1}
my_variable <- 10
my_variable

```

Fixing it makes the code run fine.

**Question 2**
```{r Q2}
library(tidyverse)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)

```

**Question 3**

Pressing Alt + Shift + K pops up keyboard shortcuts. We can get to the same place in the menu under Tools -> Keyboard Shortcuts Help.

\newpage
**Part 11.2.2**

**Question 1** 

To read a file where fields are separated with “|”, we would use the read_delim() function with the argument delim = "|". 

**Question 2** 

List of all arguments read_csv() and read_tsv() have in common,
```{r}
intersect(names(formals(read_csv)), names(formals(read_tsv)))
```

read_csv() and read_tsv() have the same arguments. 

**Question 3** 

col_positions(), which defines the column positions, is the most important argument to read_fwf(). 
