---
title: "Lecture Assignment 15"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(nycflights13)
library(tidyverse)
```


\newpage
**Part 13.4.6**

**Question 1** 

```{r Q1}
avg_delays <-flights %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

avg_delays %>%
  ggplot(aes(x = lon, y = lat, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```

\newpage
**Question 2**
```{r Q2}
locations <- airports %>%
  select(faa, lat, lon)

flights %>%
  select(year:day, hour, origin, dest) %>%
  left_join(locations, by = c("origin" = "faa")) %>%
  left_join(locations, by = c("dest" = "faa"),
    suffix = c("Origin", "Dest"))
```

**Question 5**

From June 12 to June 13, there was a series of derechos, a widespread storm, in the southeastern and Midwest US.
```{r}
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  ggplot(aes(y = lat, x = lon, size = delay, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap() +
  scale_colour_viridis_c()
```
As shown in the plot above, the largest delays were in the southeast and the Midwest, which was where the storm occurred across.

\newpage
**Part 13.5.1**

**Question 1** 

Flights having a missing tailnum means flight being cancelled, as they have msissing values of arr_time.

```{r}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(carrier, sort = TRUE)
```
Tail numbers that don’t have a matching record in planes are registered to American Airlines (AA) or Envoy Airlines (MQ).

**Question 2**

```{r}
flights100 <- flights %>%
  filter(!is.na(tailnum)) %>%
  group_by(tailnum) %>%
  count() %>%
  filter(n >= 100)

flights %>%
  semi_join(flights100, by = "tailnum")
```



