---
title: "Lecture Assignment 17"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```

\newpage

**Part 15.3.1**

**Question 1**

Exploring the distribution of **rincome** with default bar chart,

```{r Q1}
ggplot(gss_cat, aes(rincome)) +
  geom_bar()
```

With the default bar chart, the overlapping of labels on the x axis makes it hard to understand. We can improve the plot by flipping the x and y axis.

```{r}
ggplot(gss_cat, aes(rincome)) +
  geom_bar() +
  coord_flip()
```

**Question 2**

```{r Q2}
gss_cat %>%
  count(relig) %>%
  arrange(desc(n)) %>%
  head(1)
```

"Protestant" is the most common relig.

```{r}
gss_cat %>%
  count(partyid) %>%
  arrange(desc(n)) %>%
  head(1)
```

"Independent" is the most common partyid.

**Question 3**

Finding out with a table,

```{r}
levels(gss_cat$denom)
```

From this we can see that the relig, denom applies to, is "Protestant".

```{r}
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

From the scatterplot above, we can also see that the relig, denom applies to, is "Protestant".

\newpage

**Part 15.4.1**

**Question 1**

Checking the summary and visualizing the data,

```{r}
summary(gss_cat$tvhours)

gss_cat %>%
  filter(!is.na(tvhours)) %>%
  ggplot(aes(x = tvhours)) +
  geom_histogram(binwidth = 1)
```

The tv hours and the mean looks fine to me. However, the potential of mean depends on what it is used for.

**Question 2**

Factors in gss_cat,

```{r}
keep(gss_cat, is.factor) %>% 
  names()
```

For **marital**,

```{r}
levels(gss_cat$marital)

ggplot(gss_cat, aes(marital)) +
  geom_bar()
```

The order of the levels for **marital** can be principled or arbitrary. This is because the way it is ordered does make a little sense but at the same time, it is not very principled.

For **race**,

```{r}
levels(gss_cat$race)

ggplot(gss_cat, aes(race)) +
  geom_bar()
```

The order of the levels for **race** is principled as it is ordered by counts.

For **rincome**,

```{r}
levels(gss_cat$rincome)

ggplot(gss_cat, aes(rincome)) +
  geom_bar() +
  coord_flip()
```

The order of the levels for **rincome** is principled if we exclude "Not applicable", "Refused", "Don't know", and "No answer". The **rincome** is ordered in decreasing order of income.

For **partyid**,

```{r}
levels(gss_cat$partyid)

ggplot(gss_cat, aes(partyid)) +
  geom_bar() +
  coord_flip()
```

The order of the levels for **partyid** is principled as it is ordered from "Strong republican" to "Strong democrat".

For **relig**,

```{r}
levels(gss_cat$relig)

ggplot(gss_cat, aes(relig)) +
  geom_bar() +
  coord_flip()
```

The order of the levels for **relig** is arbitrary as there is no natural ordering.

For **denom**,

```{r}
levels(gss_cat$denom)

ggplot(gss_cat, aes(denom)) +
  geom_bar() +
  coord_flip()
```

The order of the levels for **denom** is also arbitrary as there is no natural ordering.
