---
title: "Lecture Assignment 11"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```


\newpage 
**Part 11.3.5**

**Question 1**

locale is used to set date and time formats, time zone, numbers, and encoding. Therefore, the most important arguments to locale() are date_names, date_format, time_format, tz, decimal_mark, grouping_mark, and encoding.


**Question 2**

Setting decimal_mark and grouping_mark to the same character throws an error stating that they both must be different.

```{r}
locale(decimal_mark = ",")
locale(grouping_mark = ".")
```

When the decimal_mark is set to ",", the default value of grouping_mark is ".". When the grouping_mark is set to ".", the default value of decimal_mark is ",".

**Question 5**

read_csv() reads comma delimited files, read_csv2() reads semicolon separated files (common in countries where , is used as the decimal place).

\newpage
**Question 7**

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"

parse_date(d1, "%B %d, %Y")
parse_date(d2, "%Y-%b-%d")
parse_date(d3, "%d-%b-%Y")
parse_date(d4, "%B %d (%Y)")
parse_date(d5, "%m/%d/%y")
parse_time(t1, "%H%M")
parse_time(t2, "%H:%M:%OS %p")
```




