---
title: "Lecture Assignment 16"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```


\newpage
**Part 14.2.5**

**Question 1** 
The paste() function by default separates strings by space, whereas, the paste0() function does not separate strings by space by default.
They are equivalent to the str_c() function, however, it is more like paste0() as it does not separate strings with spaces by default. 
While the paste() and paste0() functions convert NA to the string "NA" and treat it as a character vector, str_c() returns NA if any argument is a missing value. 


**Question 3**
```{r Q2}
x <- c("Apple", "Banana", "Pear")
len <- str_length(x)
f <- ceiling(len / 2)
str_sub(x, f, f)
```
If the string has an even number of characters, I choose ceiling(length/2) as it also accounts for cases when the length of string is one.

**Question 4**

The str_wrap() function wraps text, fitting within a certain width. We might want to use this function when wrapping long texts to be typeset.

**Part 14.3.1.1**

**Question 1**

"\" will escape the next character in R string.
"\\" will resolve to \ in regular expression, escaping the next character in the regular expression. 
For "\\\", the first 2 \\ will resolve to \ in regular expression, and the third escaping the next character.

**Question 2**

```{r}
str_view("\"'\\", "\"'\\\\", match = TRUE)
```

**Part 14.3.2.1**

**Question 1**
```{r}
str_view(c("$^$", "ab$^$sfas"), "^\\$\\^\\$$", match = TRUE)
```

**Question 2**

1) Words starting with "y", 
```{r}
str_view(stringr::words, "^y", match = TRUE)
```

2) Words ending with "x",
```{r}
str_view(stringr::words, "x$", match = TRUE)
```

3) Words that are exactly three letters long,
```{r}
str_view(stringr::words, "^...$", match = TRUE)
```

4) Words having seven letters or more,
```{r}
str_view(stringr::words, ".......", match = TRUE)
```


