---
title: "Lecture Assignment 4"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```


\newpage 
**Part 3.7.1**

**Question 1** 

The default geom associated with stat_summary() is geom_pointrange(). 
Below is how you could rewrite the previous plot to use geom_pointrange() instead of the stat function,

```{r Q1}

ggplot(data = diamonds) +
  geom_pointrange(mapping = aes(x = cut, y = depth), stat = "summary", 
                  fun.min = min, fun.max = max, fun = median)

```


**Question 2**

geom_col() is a wrapper over geom_bar() and has stat_identity() as the default stat, which is different than the default stat of geom_bar(). 
The geom_col() needs both x and y values from the data, representing the bar height. Whereas, geom_bar() only needs an x variable as it uses stat_count() as a default stat, which pre-processes input data by counting the number of observations for each value of x and so, the values of these counts are used for the y variable. 

\newpage 
**Part 3.8.1**

**Question 1**

Because there are multiple observations for each combination of cty and hwy values, there is overplotting. \newline
We could improve it by using a jitter position adjustment as it shows the area with more observations.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter")
```

**Question 2**

```{r}
?geom_jitter()
```

The "width" and "height" parameters to geom_jitter() control the amount of jittering. Width controls the amount of horizontal displacement, and height controls the amount of vertical displacement.

\newpage 
**Part 3.9.1**

**Question 1**

A stacked bar chart.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = 1, fill = clarity), show.legend = FALSE, width = 1)
```

A stacked bar chart with polar coordinates is a pie chart. We also need to add the argument theta = "y" to coord_polar() to map "y" to the angle of each section. 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = 1, fill = clarity), show.legend = FALSE, width = 1) +
  coord_polar(theta = "y")
```

**Question 2**

```{r}
?labs()
```

labs() is used for modifying axis, legend, and plot labels. It adds plot title, plot subtitle, plot caption, and axis titles.  

