---
title: "Lecture Assignment 13"
author: "Viraj Vijaywargiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```


\newpage
**Part 12.4.3**

**Question 1**

The extra and fill arguments in separate() tells what to do if there are too many values and what to do if there aren't enough values respectively.

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
```
By default, separate() drops extra values and fill fills columns with missing values, giving a warning for both. 
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")
```
With extra = "drop" or fill = "right", it produces the same results as before but without warnings. Alternatively, we can also use extra = "merge" and fill = "left". 
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")
```
With extra = "merge", the extra values merge into one so f and g becomes "f,g". With fill = "left", it fills the missing value but, from the left. 

**Question 2**

The remove() argument, in both unite() and separate(), discards the input columns in resulted data frame. In the case of creating a new variable, but keeping the old one, you would set it to FALSE. 

**Part 12.5.1**

**Question 1**

The fill arguments to pivot_wider() and complete() both are used to set values to replace NAs. While fill to pivot_wider() sets all values to replace NAs, fill to complete() does the same going through a list of names allowing for different variables and values. 

**Question 2**

The direction argument to fill() states whether NA value(s) should be replaced by the next non-missing value(s), using "up", or replaced by the previous non-missing value(s), using "down". 

\newpage
**Part 12.6.1**
```{r}
who1 <- who %>% 
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  )

who1 %>% 
  count(key)

who2 <- who1 %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))

who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")

who3 %>% 
  count(new)

who4 <- who3 %>%
  select(-new, -iso2, -iso3)

who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
```

\newpage
**Question 2**

```{r}
temp_who3 <- who1 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
```
Neglecting the mutate step results in separate() giving a warning for too few values as it expects 3 pieces. 

**Question 3**

```{r}
select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)
```
There is only one distinct combination of iso2 and iso3 values within each country, therefore, iso2 and iso3 are redundant with country. 

**Question 4**

```{r}
who5 %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(cases)) %>%
  unite(new, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = new, colour = sex)) +
  geom_line()
```





