---
title: "Stats 111 HW2"
author: "Viraj Vijaywargiya"
date: "2023-02-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(epitools)
library(rmeta)
library(pROC)
library(nnet)
```

1.  1a) The Null hypothesis will be H0 : pij = pi pj for all i=1,2,3,4 and j=1,2. That is CHD and Cholesterol level group are independent.

    Alternative Hypothesis HA: at least one combination of i and j has pij != pi pj. That is CHD and Cholesterol level group are not independent.

    1b)

    ```{r}
    chol = matrix(c(307,246,439,245,12,8,31,41),4,2)
    chisq.test(chol)
    ```

    Test statistic: 35.028, P-value: 1.202e-07.

    1c) The p-value \< 0.05, therefore, we would reject the null and conclude evidence for the\
    alternative, that CHD and Cholesterol level group are dependent. Hence, we have evidence that Cholesterol level is associated with CHD status.

    1d) The levels of cholesterol can be considered as an ordinal variable because it can have a defined order of ranking. The categories have a meaningful order and can be ranked as "Normal", "Above Normal", "High", "Very High". The fact that high cholesterol is generally accepted as leading to heart disease supports the meaningful order of the ranges.

    1e) The null hypothesis is H0 : pi1 = pi2 = pi3 = pi4 (CHD and Cholesterol level group are independent) The alternative hypothesis is Ha : pi1 \> pi2 \> pi3 \> pi4 or Ha : pi1 \< pi2 \< pi3 \< pi4 (CHD and Cholesterol level group are dependent, and there is a trend in the effect of Cholesterol level on CHD fate).

    1f)

    ```{r}
    cholesterol = array(c(307,246,439,245,12,8,31,41), dim=c(4,2), dimnames=list( Cholesterol=c("Normal","Above Normal","High","Very High"), CHD=c("no CHD","CHD")))
    n.chol = cholesterol[,2]
    n.strata = rowSums(cholesterol)
    prop.trend.test(n.chol, n.strata )
    ```

    The p-value \< 0.05, therefore, we would reject the null and conclude evidence for the\
    alternative, that CHD and Cholesterol level group are dependent. Hence, we have evidence that the probability of CHD increases or decreases as we move along the Cholesterol level group.

2.  

    ```{r}
    scout = array( c(169,43,42,11,59,196,10,10),dim = c(2, 2, 2),dimnames = list( Scout = c("No","Yes"),Verdict = c("Delinquet No", "Delinquet Yes"),Socioeconomic = c("Low","High")))
    ```

    2a) Null Hypothesis H0: Scout status and Socioeconomic level are independent.

    Alternate Hypothesis HA: Scout status and Socioeconomic level are not independent.

    Using the chi-squared test of independence,

    ```{r}
    socio = array(c(211,69,54,206), dim=c(2,2),dimnames=list(Socioeconomic=c("Low","High"), Scout=c("No","Yes")))
    chisq.test(socio)
    ```

    P-value \< 2.2e-16. Therefore, we reject the null hypothesis and conclude the alternate, that is Scout status and Socioeconomic level are not independent. Thus, Socioeconomic level is associated with Scout status.

    2b) Null Hypothesis H0: Delinquency status and Socioeconomic level are independent.

    Alternate Hypothesis HA: Delinquency status and Socioeconomic level are not independent.

    Using the chi-squared test of independence,

    ```{r}
    socio.deliq = array(c(212,255,53,20), dim=c(2,2),dimnames=list(Socioeconomic=c("Low","High"), Deliquent=c("No","Yes")))
    chisq.test(socio.deliq)
    ```

    P-value = 2.689e-05. Therefore, we reject the null hypothesis and conclude the alternate, that is Delinquency status and Socioeconomic level are not independent. Thus, Socioeconomic level is associated with Delinquency status.

    2c) Socioeconomic status can be a potential confounder in the association between Scout status and Delinquency status because it is associated with both Scout status and Delinquency status.

    2d) Mantel-Haenszel common odds ratio: 0.6570297. Therefore, the common odds ratio is not equal to 1.

    2e) The null hypothesis is H0 : OR(MH) = 1. The alternative hypothesis Ha : OR(MH) != 1.

    ```{r}
    mantelhaen.test(scout)
    ```

    P-value = 0.1843. Since the p-value \> 0.05, we fail to reject the null hypothesis.

    2f) Socioeconomic status is an effect modifier for the association between scout status and delinquency status because it changes the strength and direction of the relationship. For children from low socioeconomic backgrounds, being a scout is associated with a slightly higher odds of being delinquent (odds ratio of 1.03), while for children from high socioeconomic backgrounds, being a scout is associated with a significantly lower odds of being delinquent (odds ratio of 0.30). This indicates that the association between scout status and delinquency status is influenced by the child's socioeconomic background.

3.  3a) The estimate B0 is the intercept of the regression model, representing the expected mean systolic blood pressure (BP) when both alcohol drinking status and tobacco smoking status are equal to 0. In other words, it represents the baseline systolic blood pressure in the population for individuals who don't drink alcohol and don't smoke tobacco. This is a meaningful interpretation, as it provides a reference point for comparing the effect of alcohol drinking and tobacco smoking on systolic blood pressure.

    3b) The B3 term is the coefficient of the interaction between alcohol drinking status (X1/Drink) and tobacco smoking status (X2/Smoke). It represents the additional effect of alcohol drinking on systolic blood pressure that is specific to individuals who smoke tobacco. The B3 term provides information on how the association between alcohol drinking and systolic blood pressure is modified by tobacco smoking status. For example, if B3 is positive and significantly different from zero, it suggests that the effect of alcohol drinking on systolic blood pressure is stronger for individuals who smoke tobacco than for those who don't.

    3c) For non-smokers (X2/Smoke = 0), the expected systolic blood pressure for a person who drinks (X1/Drink = 1) can be calculated as BP = B0\^ + B1\^ \*1 + B2\^ \*0 + B3\^ \*1 \*0 + E = B0\^ + B1\^ + E. This equation suggests that the effect of drinking on systolic blood pressure for non-smokers is given by B1\^.

    For smokers (X2/Smoke = 1), the expected systolic blood pressure for a person who drinks (X1/Drink = 1) can be calculated as BP = B0\^ + B1\^ \*1 + B2\^ \*1 + B3\^ \*1 \*1 + E = B0\^ + B1\^ + B2\^ + B3\^ + E. This equation suggests that the effect of drinking on systolic blood pressure for smokers is given by B1\^ + B2\^ + B3\^.

    In conclusion, the effect of drinking on systolic blood pressure is different for smokers and non-smokers and is given by B1\^ + B2\^ + B3\^ for smokers and B1\^ for non-smokers.

    3d) Null hypothesis H0: B2 = B3 = 0. The effect of tobacco smoking (X2/Smoke) on systolic blood pressure (BP) is equal to zero, both as a main effect and as an interaction with alcohol drinking status (X1/Drink).

    Alternate hypothesis HA: B2 != 0 or B3 != 0. The effect of tobacco smoking (X2/Smoke) on systolic blood pressure (BP) is not equal to zero, either as a main effect or as an interaction with alcohol drinking status (X1/Drink).

    Model 1: BP(i) = B0 + B1Drink(i) + B2Smoke(i) + E(i)

    Model 2: BP(i) = B0 + B1Drink(i) + B2Smoke(i) + B3Drink(i)Smoke(i) + E(i)

4.  

    ```{r}
    framingham = read.table("/Users/virajvijaywargiya/Downloads/framingham.txt")

    framingham$sex = framingham$sex - 1
    names( framingham )[1] = "female"
    framingham[1:10,]

    framingham$sbphi = cut( framingham$sbp, breaks=c(min(framingham$sbp),146, max(framingham$sbp)), include.lowest=TRUE )

    ```

    4a) BMI is a possible cofounder in the relationship between systolic blood pressure and the prevalence of CHD because it could independently affect both systolic blood pressure and CHD. A higher BMI is associated with both higher blood pressure and increased risk of CHD, so if not controlled for, it could artificially inflate the observed association between systolic blood pressure and CHD.

    4b)

    ```{r}
    framingham$bmigrp = cut( framingham$bmi, breaks=c(min(framingham$bmi, na.rm=TRUE),20, 25, 30, max(framingham$bmi, na.rm=TRUE)), include.lowest=TRUE, right=FALSE )
    ```

    Test of independence and test for trend for association between BMI and SBP,

    ```{r}
    bmisbp.table = xtabs( ~ bmigrp + sbphi, data=framingham )
    epitab( bmisbp.table, pvalue="chi2" )
    n.hisbp = bmisbp.table[,2]
    n.strata = rowSums(bmisbp.table) 
    chisq.test(bmisbp.table)
    prop.trend.test( n.hisbp, n.strata )
    ```

    Since the p-value \< 0.05, we reject the null hypothesis (BMI and SBP are independent) and conclude the alternate (BMI and SBP are not independent). Therefore, BMI is associated with systolic blood pressure (SBP).

    Test of independence and test for trend for BMI and CHD,

    ```{r}
    bmichd.table = xtabs( ~ bmigrp + chdfate, data=framingham )
    epitab( bmichd.table, pvalue="chi2" )
    n.chd = bmichd.table[,2]
    n.strata = rowSums(bmichd.table) 
    chisq.test(bmichd.table)
    prop.trend.test( n.chd, n.strata )
    ```

    Since the p-value \< 0.05, we reject the null hypothesis (BMI and CHD are independent) and conclude the alternate (BMI and CHD are not independent). Therefore, BMI is associated with CHD.

    Therefore, the data suggest that your investigators are correct in their prior belief that BMI is a potential cofounder in the relationship between systolic blood pressure and the prevalence of CHD because BMI is associated with both systolic blood pressure and the prevalence of CHD.

    4c)

    ```{r}
    n.sbplo = xtabs( ~ sbphi + bmigrp, data=framingham )[1,]
    n.sbphi = xtabs( ~ sbphi + bmigrp, data=framingham )[2,]
    lo.chd = xtabs( ~ chdfate + sbphi + bmigrp, data=framingham )[2,1,]
    hi.chd = xtabs( ~ chdfate + sbphi + bmigrp, data=framingham )[2,2,]
    ```

    Odds ratio and corresponding CI for association between BMI and SBP,

    BMI 20-25: OR = 1.62, CI = [1.04, 2.51]

    BMI 25-30: OR = 3.21, CI = [2.08, 4.95]

    BMI \>=30: OR = 6.62, CI = [4.22, 10.4]

    Odds ratio and corresponding CI for association between BMI and CHD,

    BMI 20-25: OR = 2.16, CI = [1.49, 3.13]

    BMI 25-30: OR = 3.65, CI = [2.53, 5.28]

    BMI \>=30: OR = 4.41, CI = [2.98, 6.53]

    4d)

    ```{r}
    mh.rslt = meta.MH( n.sbphi, n.sbplo, hi.chd, lo.chd, names=levels(framingham$bmigrp) )
    summary( mh.rslt )
    ```

    The Mantel-Haenszel estimate of the common odds ratio = 1.66 with CI = [1.43, 1.92]

    4e)

    ```{r}
    framingham$bmi_categories = cut(framingham$bmi,
      breaks = c(min(framingham$bmi, na.rm=TRUE) - 1, 20, 25, 30, max(framingham$bmi, na.rm=TRUE)+1), include.lowest=TRUE, right=FALSE,
      labels = c("Low", "Normal", "High", "Obese")) 

    framingham$sbphi = cut( framingham$sbp, breaks=c(min(framingham$sbp),146, max(framingham$sbp)), include.lowest=TRUE )

    ntrt = table(framingham$sbphi, framingham$bmi_categories)[2,]
    nctrl = table(framingham$sbphi, framingham$bmi_categories)[1,]
    ptrt = table(framingham$sbphi, framingham$bmi_categories, framingham$chdfate)[2,,2]
    pctrl = table(framingham$sbphi, framingham$bmi_categories, framingham$chdfate)[1,,2]

    bd.test = meta.MH(ntrt, nctrl, ptrt, pctrl)
    bd.test
    summary(bd.test) 
    ```

    The null is H0 : OR1 = OR2. Alternate HA: Null is not true.

    Since the p-value \> 0.05, Breslow-Day test fails to reject the null that all the odds ratios are equal, which opens us up to use the MH test and MH common odds ratio.\
