---
title: "Stats 110 HW2"
author: "Viraj Vijaywargiya"
date: "2022-10-17"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
```

**1)1a)** False. **1b)** False. **1c)** True. **1d)** False. **1e)**
True. **1f)** False.

**2)2a)** Same, as correlation between X and Y is the same as the
correlation between Y and X. **2b)** Same, as R is the same, R-squared
will be the same. **2c)** Different, as B1 is the slope and as Y and X
change, B1 will differ. **2d)** Different, same reason as part c.
**2e)** Same, as test statistic depends only on R and since R is the
same, the test statistic will be the same. **2f)** Same, as residuals
for each observation stays the same as the relation between X and Y
stays the same.

**3)**

```{r}
# Import MidWestSales dataset
    MidWestSales = read.table("/Users/virajvijaywargiya/Downloads/MidwestSales.txt", fill=TRUE, header=FALSE)
    # This dataset dsigma(e)s not have names, so we will add names to the variables
    names(MidWestSales)=c("id","price","sqft","bed","bath","ac","garage","pool","year","quality","style","lot","hwy")
    view(MidWestSales)
```

**3a)**

```{r}
lm(formula = price ~ sqft, data = MidWestSales)
```

Estimated regression equation: Y = -81433 + 159X.

**3b)** As the square footage of the house increases by 1 unit, the
expected sale price increases by 159 dollars.

**3c)**

```{r}
model = lm(formula = price ~ sqft, data = MidWestSales)
summary(model)
n = dim(MidWestSales)[1]
r = cor(MidWestSales$price, MidWestSales$sqft)
Rsquared = summary(model)$r.squared
t = (r*sqrt(n-2))/(sqrt(1-Rsquared))
p = 2*(1-pt(abs(t),n-2))
```

Null hypothesis: B1 = 0, Alternate hypothesis: B1 != 0. Test statistic:
t = 32.605, Two sided p-value = 2.2e-16. Since the p-value is
essentially 0, which is less that the a = 0.05 level of significance.
Therefore, we reject the null hypothesis and can conclude that sqft has
a linear relationship with price (alternate hypothesis).

**3d)** Null hypothesis: B1 = 0, Alternate hypothesis: B1 \> 0. For
one-sided p-value we divide it by 2, 2.2e-16/2, which is also
essentially 0. B1 \> 0 and the t\* is also positive. Therefore, we
reject the null hypothesis and conclude that sqft has a positive linear
relationship with price (alternate hypothesis).

**3e)**

```{r}
predict(model, list(sqft=2000),interval= "c")
```

We are 95% confident that the interval (229220.7, 243714.4) contains the
true mean of the response (sale price) at X=2000 (2000 sqft).

**3f)**

```{r}
predict(model, list(sqft=2000),interval= "p")
```

We are 95% confident that the interval (80858.85, 392076.2) contains the
true response (sale price) at X=2000 (2000 sqft).

**3g)**

```{r}
predict(model, list(sqft=2000),interval= "p", level = 0.90)
```

**3h)** No, it wouldn't make sense as 8500 is greater than the maximum
value for sqft. As the model is not fit for that large of a value,
predicting the price of a 8500 sqft house may lead to inaccuracy.

**3i)**

```{r}
predict(model, list(sqft=2000),interval= "p", level = 1)
```

The lower and upper limit becomes -infinite and infinite respectively.
This is because a 100% confidence means that there is no doubt at all
that we will get the same results if the study is repeated.

**3j)** The estimate of the sigma(e) is 79120, which means that the
standard deviation for the approximate distribution of Y is 79120.

**3k)** Yes it dsigma(e)s make sense for both these houses to have the
same estimate of sigma since the standard deviation (or variance) of the
Y values is constant for all values of X in the range of the data.

**4)**

```{r}
skincancer = read.table("/Users/virajvijaywargiya/Downloads/skincancer.txt", fill = TRUE, header = TRUE)
NewModel = lm(formula = Mort ~ Lat, data = skincancer)
summary(NewModel)
```

**4a)**H0 : B1 = 0 vs Ha : B1 6= 0. Test statistic is t\* = -9.99. Two
sided p-value is 3e-13 which is essentially 0. Therefore, reject the
null and conclude evidence for B1 \> 0.

**4b**)

```{r}
predict(NewModel, list(Lat=40),interval= "c", level = 0.99)
```

The 99% confidence interval for the true mean of the response variable
when Lat=40 is (142.7148, 157.453). This means we are 99% confident that
the true average mortality at 40 Lat is within that range.

**4c)**

```{r}
predict(NewModel, list(Lat=40),interval= "p", level = 0.99)
```

The 99% confidence interval for the true mean of the response variable
when Lat=40 is (98.24214, 201.9257). This means we are 99% confident
that the true mortality at 40 lat is within that range.

**4d)** Both use the same estimate of their respective parameters, uY
for mean and Y for true value.

**4e)** The prediction interval is wider than the confidence interval
because it accounts for the error term to the true mean. This will
always result in greater variances.

**5)5a)** Narrower

**5b)** Narrower

**5c)** Narrower

**5d)** Stays the same

**6)** R-squared \~ 0.73, which means that around 73% of the variance in
Y is explained/predicted by X.

**7)**

```{r}
pulse = read.table("/Users/virajvijaywargiya/Downloads/Pulse.txt", fill = TRUE, header = TRUE)
pulse$Smoker = ifelse(pulse$Smoke==1, "Yes", "No")
view(pulse)
```

**7a)**

```{r}
pulsemodel = lm(formula = Rest ~ Smoker, data = pulse)
summary(pulsemodel)
```

Null hypothesis: B1 = 0, Alternate hypothesis: B1 != 0. Test statistic:
t-value = 2.429, F-statistic = 5.902. p-value = 0.01589. Therefore, we
reject the null hypothesis and conclude that the csigma(e)fficient on X2
(the smoking status) is not 0 (alternate hypothesis). This also means
that there is a relationship between resting pulse rate and smoking
status.

**7b)**

```{r}
t.test(pulse$Rest[pulse$Smoker=="Yes"], pulse$Rest[pulse$Smoker=="No"], var.equal=TRUE) 
```

Having done the two sample t-test, the t-value 2.429 with degrees of
freedom 230, and the p-value 0.01589 are the same compared to the test
from part a.

**7c)** Yi = B0 + B1X1 + B2X2 + e

**7d)** With respect to the effect of weight on resting pulse rate, not
having an interaction term means that weight dsigma(e)s not have an
effect on smoking status.

**7e)**

```{r}
summary(lm(formula = Rest ~ Wgt + Smoker, data = pulse))

```

Equation: Y = 78.25 - 0.067 X1 + 6.04X2 Csigma(e)fficient of
determination (multiple R-squared value) = 0.06978, estimate of sigma(e)
(RSE) = 9.638.

**7f)** 232

**7g)** Null hypothesis: B2 = 0, Alternate hypothesis: B2 != 0. Test
statistic: t-value = 2.975. p-value = 0.00325. Therefore, we reject the
null hypothesis and conclude that the csigma(e)fficient on smoking yes
(X2) in not 0, which means that with respect of the effect of weight on
resting pulse rate, there is a relation between resting pulse and
smoking status.

**7h)** No, we cannot conclude that smoking causes lower resting pulse
rates because it is an observational study and the cofounding variables
are not in control. Also, smoking status and resting pulse rates have a
positive relation.

**7i)**

```{r}
library(ggplot2)
pulse$Smoker = ifelse(pulse$Smoke==1, "Yes", "No")
ggplot(pulse, aes(Wgt, Rest, color = Smoker, shape=Smoker)) + geom_point()+geom_smooth(method="lm", se=FALSE)
```

**7j)** Yi = B0 + B1X1 + B2X2 + (BX1X2 )+ e

**7k)**

```{r}
summary(lm(formula = Rest ~ Wgt + Smoker + Wgt*Smoker, data = pulse))
```

Regression equation: Y = 80.43 - 0.081 X1 - 10.031 X2 + (0.095 X1X2)

**7l)** R-squared value = 0.08105, sigma(e) = 9.6. Compared to part e,
there is a slight increase in the R-squared value, but the sigma(e)
remains about the same.

**7m)** Null hypothesis: B3 = 0, Alternate hypothesis: B3 != 0. Test
statistic: t-value = 1.672. p-value = 0.095863. Therefore, we fail to
reject the null hypothesis, and can conclude that the effect of weight
on resting pulse rate can differ for smokers and non-smokers.

**7n)** No, it is not a good fitting model based on the R-squared value
because a value of 0.07 means that only 7% of the variance in Y is
explained by the eplanatory variables Xi.

**7o)** Exercising increases blood flow to the hear and also increases
the pulse rates. This could be a cofounder in this study as the the
response variable pulse rate dsigma(e)s get affected by Exercise amount.

**7p)** This can be done by checking if the r-squared value increases
when changing X1. Null hypothesis: X1 \> 0, Alternate hypothesis: X1 \<=
0.
